KiCS2: A New Approach to Compile Curry to Haskell

### 2011-03-30 papers/wflp2011/paper.tex

\newcommand{\code}[1]{\mbox{\texttt{#1}}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92\xspace} % backslash
\newcommand{\us}{\char95\xspace} % underscore
\newcommand{\Choice}[2]{#1 \mathop{?} #2}
\newcommand{\ol}[1]{\overline{#1}}

@@

\newcommand{\Cc}{{\cal{C}}} % constructors
\newcommand{\Fc}{{\cal{F}}} % defined functions
\newcommand{\Var}{{\cal V}ar} % variables in an object
\newcommand{\Dom}{{\cal D}om} % domain of a substitution
\newcommand{\Rc}{{\cal{R}}} % rewrite system
\newcommand{\Tc}{{\cal{T}}} % algebra of terms
\newcommand{\Xc}{{\cal{X}}} % set of all variables
\renewcommand{\emptyset}{\varnothing}
\newcommand{\suc}{\mathit{success}}  % success
\newcommand{\Suc}{\mathtt{Success}}  % success type
\newcommand{\funset}{\ensuremath{_{\cal S}}}
\newcommand{\seq}{\mathrel{\code{=:=}}} % strict equality of math mode

@@

\begin{abstract}
bla bla
\end{abstract}

@@

In this paper we describe an implementation that is based
on the latter principles. We present a method to compile
programs written in the functional logic language Curry \cite{Hanus06Curry}
into Haskell programs
The difficulty of such an implementation is the fact
that nondeterministic results can occur in any place of a
computation. Thus, one cannot separate logic computations
by the use of list comprehensions \cite{Wadler85},
but any outcome of some operation could be potentially nondeterministic,
i.e., it might have more than one result value.
We solve this problem by an explicit representation of
nondeterministic values, i.e., we extend each data type by
another constructor to represent the choice between several values.
This idea is also the basis of Curry implementation KiCS
\cite{BrasselHuch07,BrasselHuch09}.
However, KiCS is based on unsafe features of Haskell
that inhibit the use of optimizations provided by Haskell compilers
like GHC.\footnote{\url{http://www.haskell.org/ghc/}}
In contrast, our implementation is not based on unsafe features.
In addition, we also support more flexible search strategies
and new features to encapsulate nondeterministic computations.

@@

In the next section, we sketch are source language Curry
and introduce a normalized form of Curry programs that is the
basis of our translation scheme.
Section~\ref{sec:Compilation} presents the basic ideas
of this translation scheme.
Further aspects related to our compiler are discussed in
Section~\ref{sec:FurtherAspects}.
We show benchmarks of our initial implementation in
Section~\ref{sec:Benchmarks}
before we conclude in Section~\ref{sec:Conclusions}.

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\
    &  |  & \mbox{\code{let}}~\{x_1=e_1;\ldots;x_n = e_n\}~\code{in}~e
                                            & \mbox{let binding}\\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & \ol{D_k}
\end{array}
\]
\caption{Uniform programs}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.
There, %$\ol{o_n}$ denotes the object sequence $o_1,\ldots,o_n$,
the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules where only one constructor occurs in the same argument
of all rules.\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Brassel \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following sections.

@@

We can correct this wrong behavior of our implementation
by identifying different \code{Choice} occurrences that our duplicates
of some single \code{Choice}. This can be easily done by attaching
a unique identifier, e.g., a number, to each choice:
\begin{haskell}
  type ID = Integer
  data Bool = False | True | Choice ID Bool Bool
\end{haskell}
Furthermore, we modify the \code{Choice} pattern rules so that
the identifiers will be kept, e.g.,
\begin{curry}
  xor (Choice i x1 x2) x = Choice i (xor x1 x) (xor x2 x)
\end{curry}
If the expression \code{aBool} assigns the number \code{1}
to its choice, we obtain the result
\begin{lstlisting}
  Choice 1 (Choice 1 False True) (Choice 1 True False)
\end{lstlisting}
when evaluating the expression \ccode{xorSelf aBool}.
If we show the values contained in this result,
we have to make \emph{consistent} selections in choices with
same identifiers. This, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 1 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 1 True False)} which yields to only
value \code{False}.

@@

\code{thisID} creates a fresh identifier from a current
identifier supply, and
\code{leftSupply} and \code{rightSupply} create
new distinct identifier supplies from a given one.
The latter are used when an operation based on two\footnote{The
extension to more than two is straightforward.}
other operations in the right-hand side of a rule. In this case,
the other operations must be supplied with their individual distinct
identifier supplies. For instance, the operation \code{main}
defined by
\label{sec:xor-main}
\begin{curry}
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
Any choice in the right-hand side of a rule gets its own identifier
by the operation \code{thisID}, as in
\begin{haskell}
  aBool s = Choice (thisID s) True False
\end{haskell}
There are various implementations of \code{IDSupply}.
The simplest implementation uses integers:
\begin{haskell}
  type IDSupply = Integer
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
  thisID      n = n
\end{haskell}
Actually, our compilation system is parameterized over different
implementations of \code{IDSupply}
in order to perform some experiments and choose
he most appropiate for a given application.
Each implementation must ensure that, if $s$ is a value of type
\code{IDSupply}, then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
are different identifiers provided that
$o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$
and $o_1\cdots o_n \neq o'_1\cdots o'_m$.

@@

So far, our generated operations compute all the nondeterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail = Fail
  try x = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation support various options together with various
implementations of \code{IDSupply}.
For instance, a simple implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are just memory cells in Haskell:\footnote{%
In this case, integers cannot be used for \code{IDSupply}.
Nevertheless,
this type can be implemented as an infinite tree containing
all memory cells required in a computation that are created as
needed with initial value \code{NoChoice}.}
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
The implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

\begin{figure}
\centering
\begin{tabular}{|l|r|r|r|r|}
\hline
System  & Reverse & ReversePrim &  Tak  & TakPeano \\
\hline
KiCS2   &    0.13 &        0.11 &  0.24 &     0.82 \\
 \hline
PAKCS   &    2.07 &        1.91 & 39.19 &    60.69 \\
\hline
MCC     &    0.37 &        0.38 &  1.03 &     3.99 \\
\hline
GHC     &    0.11 &        0.10 &  0.04 &     0.50 \\
\hline
SICStus &    0.39 &        0.28 &  0.50 &     4.82 \\
\hline
SWI     &    1.33 &        1.14 &  1.87 &    12.72 \\
\hline
\end{tabular}
\caption{Benchmarks: first-order functional programs}
 \label{fig:bench-first-order}
\end{figure}
%
The first collection of benchmarks (Fig.~\ref{fig:bench-first-order})
are purely first-order functional programs.
The Prolog (SICStus, SWI) and Haskell (GHC) programs have been rewritten
according to the Curry formulation.
``Reverse'' is the naive reverse program applied to a list of 4096 elements.
``ReversePrim'' is the same but with user-defined lists instead of
the predefined lists. ``Tak'' is highly recursive tak function applied
to arguments (27,16,8) and ``TakPeano'' is the same but one
user-defined natural numbers in Peano representation.
Note that the Prolog programs uses a strict evaluation strategy
in contrast to all others. Thus, the difference between PAKCS and SICStus
shows the overhead to implement lazy evaluation in Prolog.

- higher-order

- nondeterministic

- logic variables

- sharing over nondet for free!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Further Aspects}
\label{sec:FurtherAspects}

- discuss search strategies

- encapsulated search

- free variables and unification

- set functions??

@@

better program analysis to approximate purely functional and
higher-order computations

>>>>

\newcommand{\code}[1]{\mbox{\small\texttt{#1}}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92\xspace} % backslash
\newcommand{\us}{\char95\xspace} % underscore
\newcommand{\Choice}[2]{#1 \mathop{?} #2}
\newcommand{\ol}[1]{\overline{#1}}

@@

\newcommand{\funset}{\ensuremath{_{\cal S}}}

@@

\begin{abstract}
In this paper we present our first results of a new system
to compile functional logic programs of the source language Curry
into purely functional Haskell programs.
Our implementation is based on the idea to represent
nondeterministic result values of an expression as a data structure.
This enables the application of various search strategies
to extract the different values from the search space.
In addition to the increased functionality of our system,
we demonstrate by a several benchmarks that our implementation
can compete with or outperforms other existing implementations of Curry.
\end{abstract}

@@

In this paper we describe an implementation that is based
on the latter principles. We present a method to compile
programs written in the functional logic language Curry \cite{Hanus06Curry}
into Haskell programs
The difficulty of such an implementation is the fact
that nondeterministic results can occur in any place of a
computation. Thus, one cannot separate logic computations
by the use of list comprehensions \cite{Wadler85},
but any outcome of some operation could be potentially nondeterministic,
i.e., it might have more than one result value.
We solve this problem by an explicit representation of
nondeterministic values, i.e., we extend each data type by
another constructor to represent the choice between several values.
This idea is also the basis of the Curry implementation KiCS
\cite{BrasselHuch07,BrasselHuch09}.
However, KiCS is based on unsafe features of Haskell
that inhibit the use of optimizations provided by Haskell compilers
like GHC.\footnote{\url{http://www.haskell.org/ghc/}}
In contrast, our implementation, called KiCS2, is not based on unsafe features.
In addition, we also support more flexible search strategies
and new features to encapsulate nondeterministic computations.

@@

In the next section, we sketch are source language Curry
and introduce a normalized form of Curry programs that is the
basis of our translation scheme.
Section~\ref{sec:Compilation} presents the basic ideas
of this translation scheme.
Benchmarks of our initial implementation of this scheme
are presented in Section~\ref{sec:Benchmarks}.
Further features of our system are sketched in
Section~\ref{sec:FurtherAspects}
before we conclude in Section~\ref{sec:Conclusions}.

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
%    &  |  & \mbox{\code{let}}~\{x_1=e_1;\ldots;x_n = e_n\}~\code{in}~e
%                                            & \mbox{let binding}\\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & \ol{D_k}
\end{array}
\]
\caption{Uniform programs}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.
There, %$\ol{o_n}$ denotes the object sequence $o_1,\ldots,o_n$,
the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules where only one constructor occurs in the same argument
of all rules.\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Brassel \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following sections.

@@

We can correct this unintended behavior of our implementation
by identifying different \code{Choice} occurrences that are duplicates
of some single \code{Choice}. This can be easily done by attaching
a unique identifier, e.g., a number, to each choice:
\begin{haskell}
  type ID = Integer
  data Bool = False | True | Choice ID Bool Bool
\end{haskell}
Furthermore, we modify the \code{Choice} pattern rules so that
the identifiers will be kept, e.g.,
\begin{curry}
  xor (Choice i x1 x2) x = Choice i (xor x1 x) (xor x2 x)
\end{curry}
If the expression \code{aBool} assigns the number \code{1}
to its choice, we obtain the result
\begin{lstlisting}
  Choice 1 (Choice 1 False True) (Choice 1 True False)
\end{lstlisting}
when evaluating the expression \ccode{xorSelf aBool}.
If we show the values contained in this result,
we have to make \emph{consistent} selections in choices with
same identifiers. This, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 1 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 1 True False)} which yields to only
value \code{False}.

@@

\code{thisID} creates a fresh identifier from a current
identifier supply, and
\code{leftSupply} and \code{rightSupply} create
new distinct identifier supplies from a given one.
The latter are used when an operation based on two\footnote{The
extension to more than two is straightforward.}
other operations in the right-hand side of a rule. In this case,
the other operations must be supplied with their individual distinct
identifier supplies. For instance, the operation \code{main}
defined by
\label{sec:xor-main}
\begin{curry}
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
Any choice in the right-hand side of a rule gets its own identifier
by the operation \code{thisID}, as in
\begin{haskell}
  aBool s = Choice (thisID s) True False
\end{haskell}
There are various implementations of \code{IDSupply}.
The simplest implementation uses unbounded integers:
\begin{haskell}
  type IDSupply = Integer
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
  thisID      n = n
\end{haskell}
There are other more sophisticated implementations available
\cite{AugustssonRittriSynek94}.
Actually, our compilation system is parameterized over different
implementations of \code{IDSupply}
in order to perform some experiments and choose
the most appropiate for a given application.
Each implementation must ensure that, if $s$ is a value of type
\code{IDSupply}, then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
are different identifiers provided that
$o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$
and $o_1\cdots o_n \neq o'_1\cdots o'_m$.

@@

So far, our generated operations compute all the nondeterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail = Fail
  try x = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation support various options together with various
implementations of \code{IDSupply}.
For instance, a simple implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are just memory cells in Haskell:\footnote{%
In this case, integers cannot be used for \code{IDSupply}.
Nevertheless,
this type can be implemented as an infinite tree containing
all memory cells required in a computation that are created as
needed with initial value \code{NoChoice}.}
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
The implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{4}{@{~~}r@{~~}|}}
\hline
System  & ReverseUser & Reverse &  Tak  & TakPeano \\\hline
KiCS2   &        0.13 &    0.11 &  0.24 &     0.82 \\
PAKCS   &        2.07 &    1.91 & 39.19 &    60.69 \\
MCC     &        0.37 &    0.38 &  1.03 &     3.99 \\
GHC     &        0.11 &    0.10 &  0.04 &     0.50 \\
SICStus &        0.39 &    0.28 &  0.50 &     4.82 \\
SWI     &        1.33 &    1.14 &  1.87 &    12.72 \\
\hline
\end{tabular}
\caption{Benchmarks: first-order functional programs}
 \label{fig:bench-first-order}
\end{figure}
%
The first collection of benchmarks (Fig.~\ref{fig:bench-first-order})
are purely first-order functional programs.
The Prolog (SICStus, SWI) and Haskell (GHC) programs have been rewritten
according to the Curry formulation.
``ReverseUser'' is the naive reverse program applied to a list of 4096 elements,
where all data (lists, numbers) are user-defined.
``Reverse'' is the same but with built-in lists.
``Tak'' is a highly recursive function on naturals applied
to arguments (27,16,8) and ``TakPeano'' is the same but with
user-defined natural numbers in Peano representation.
Note that the Prolog programs uses a strict evaluation strategy
in contrast to all others. Thus, the difference between PAKCS and SICStus
shows the overhead to implement lazy evaluation in Prolog.

One can deduce from these results that one of the initial
goals for this compiler is satisfied since functional Curry programs
are executed almost with the same speed as their Haskell equivalents.
An overhead is visible if one uses built-in numbers
(due to the potential nondeterministic values, KiCS2 cannot
directly map operations on numbers directly into the Haskell primitives)
where the GHC can apply particular optimizations.

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{5}{@{~~}r@{~~}|}}
\hline
System  & ReverseHO & Primes & PrimesPeano & Queens & QueensUser \\\hline
KiCS2   &     32.82 &   1.12 &        0.29 &   9.77 &     12.41  \\
KiCS2HO &      0.21 &   0.09 &        0.27 &   0.63 &      0.72  \\
PAKCS   &      7.24 &  13.55 &       22.74 &  78.52 &     81.98  \\
MCC     &      0.22 &   0.32 &        1.26 &   2.65 &      2.93  \\
GHC     &      0.21 &   0.06 &        0.21 &   0.05 &      0.12  \\
\hline
\end{tabular}
\caption{Benchmarks: higher-order functional programs}
 \label{fig:bench-higher-order}
\end{figure}
%
The next collection of benchmarks (Fig.~\ref{fig:bench-higher-order})
considers higher-order functional programs so that we drop
the comparison to first-order Prolog systems.
``Reverse'' reverses a list in linear time using higher-order functions
like \code{foldl} and \code{flip} applied to a list with one million
elements.
``Primes'' computes the the 2000.\ prime number via the
sieve of Eratosthenes using higher-order functions,
and ``PrimesPeano'' computes the 256.\ prime number but
with Peano numbers and user-defined lists.
Finally, ``Queens'' (and ``QueensUser'' with user-defined lists)
computes the number of safe positions of queens on a $11 \times 11$
chess board.

As discussed above, our compiler performs an optimization
when all operations are deterministic. However, in the case
of higher-order functions, this optimization cannot be performed
since any operation, i.e., also a nondeterministic operation,
can be passed as an argument. As shown in the first line of this table,
this considerably reduces the overall performance.
To improve this situation, our compiler generates two versions
of a higher-order function: a general version applicable to any
argument and a specialized version where all higher-order arguments
are assumed to be deterministic operations.
Moreover, we implemented a program analysis
to approximate those operations that call higher-order functions
with deterministic operations so that one can use their
specialized versions. The result of this improvement
is shown as ``KiCS2HO'' and demonstrates its usefulness so that
it is always used in the subsequent benchmarks.

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
\hline
System  & PermSort & PermSortPeano &  Last \\\hline
KiCS2   &     3.23 &          1.26 &  0.26 \\
PAKCS   &    26.08 &         21.75 &  2.58 \\
MCC     &     1.23 &          1.28 &  0.08 \\
\hline
\end{tabular}
\caption{Benchmarks: nondeterministic functional logic programs}
 \label{fig:bench-nondet}
\end{figure}
%
To evaluate the efficiency of nondeterministic computations
(Fig.~\ref{fig:bench-nondet}),
we sorted a list containing 15 elements by enumerating all permutations
and selecting the sorted ones (``PermSort'' and ``PermSortPeano''
for Peano numbers and user-defined lists with 14 elements)
and computed the last element \code{x}
of a list \code{xs} containing 100,000 elements
by solving the equation \ccode{ys++[x] =:= xs}
(the implementation of unification and variable bindings
require some additional machinery that is sketched in
Section~\ref{sec:unification}).
Although the results shows that our implementation cannot compete
with MCC, it is superior to PAKCS which exploits Prolog features
like backtracking and logic variables for these benchmarks.

Since our implementation represents nondeterministic values
as Haskell data structures, we get, in contrast to most
other implementations of Curry, one interesting improvement for free:
deterministic subcomputations are shared even if they occur
in different nondeterministic computations.
To show this effect of our implementation, consider the
nondeterministic sort operation \code{psort} (permutation sort)
and the infinite list of all prime numbers \code{primes},
as used in the previous benchmarks, and the following definitions:
\begin{curry}
  goal1 = [primes!!803, primes!!802, primes!!801, primes!!800]
  goal2 = psort [6173,6163,6151,6143]
  goal3 = psort [primes!!803, primes!!802, primes!!801, primes!!800]
\end{curry}
In principle, one would expect that the sum of the execution times
of \code{goal1} and \code{goal2} is equal to the time to execute
\code{goal3}. However, implementations based on backtracking
evaluate the primes occurring in \code{goal3} multiple times,
as can be seen by the run times for PAKCS and MCC shown
in Fig~\ref{fig:bench-sharing-over-nondet}.
%
\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
\hline
System  &   goal1  &   goal2 & goal3 \\\hline
KiCS2   &    0.32  &   0.00  &  0.32 \\
PAKCS   &    8.88  &   0.02  & 91.68 \\
MCC     &    0.20  &   0.00  &  1.94 \\
\hline
\end{tabular}
\caption{Benchmarks: sharing over nondeterminism}
 \label{fig:bench-sharing-over-nondet}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Further Features}
\label{sec:FurtherAspects}

In this section we sketch some additional features of our implementation.
Due to lack of space, we cannot discuss all details of the implementation.

\subsection{Search Strategies}

Due to the fact that we represent nondeterministic results
in a data structure rather as a computation as in implementations
based on backtracking, we can provide different methods to
explore the search space containing the different result values.
We have already seen in Section~\ref{sec:printValsDFS}
how we can explore this search space to print all values in
depth-first order.
Apart from this simple approach,
our implementation contains various strategies
(depth-first, breadth-first, iterative deepening, parallel search)
to transform a choice structure into a list of results
that can be printed in different ways (e.g., all solutions, only the first
solutions, or one after another by user requests).
Actually, the user can set options to select the search strategy and
the printing method.

\begin{figure}
\centering
\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
\hline
Method               & PermSort & PermSortPeano &  NDNums \\\hline
\code{printValsDFS}  &    3.24  &         1.26  & $\infty$ \\
depth-first search   &    3.63  &         1.40  & $\infty$ \\
breadth-first search &   18.30  &         6.85  &    11.52 \\
iterative deepening  &    5.94  &         1.20  &     3.22 \\
%parallel (1 proc.)   &    5.84  &         2.09  & $\infty$ \\
%parallel (2 proc.)   &    3.46  &         1.25  & $\infty$ \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different search strategies}
 \label{fig:search-strategies}
\end{figure}
%
In order to compare the various search strategies,
Fig.~\ref{fig:search-strategies} contains corresponding
benchmarks where ``PermSort'' and ``PermSortPeano'' are the programs
discussed in Fig.~\ref{fig:bench-nondet} and ``NDNums'' is the program
\begin{curry}
  f n = f (n+1) ? n
\end{curry}
where we look for the first solution of \ccode{f 0 == 25000}
(obviously, depth-first search strategies do not terminate on this
expression).
All strategies except for the ``direct print'' method \code{printValsDFS}
translate choice structures into monadic list structures
in order to print them according to the user options.
The benchmarks show that the overhead of this transformation
is negligible so that this more flexible approach is the default one.

We also made initial benchmarks with the parallel strategy
where nondeterministic choices are explored via GHC's \code{par}
construct. For the permutation sort we obtained a speedup of 1.7
when executing the same program on two processors, but no essential
speedup is obtained for more than two processors. 
Better results require a careful analysis of the synchronization
caused by the global structure to manage the state of choices
which will be done in future work.

\subsection{Encapsulated Search}

Instead of choosing different search strategies to evaluate
top-level expressions, our system also contains a primitive operation
\begin{curry}
 searchTree :: a -> SearchTree a
\end{curry}
to translate the search space caused by the evaluation of its argument
into tree structure of the form
\begin{curry}
 data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
\end{curry}
so that the programmer can define its own search strategy or to
collect all nondeterministic values into a list structure
for further processing.

% TODO: set functions??

\subsection{Logic Variables and Unification}
\label{sec:unification}

Although our implementation is based on eliminating all logic
variables from the source program by introducing generators,
many functional logic programs contain equational constraints
(e.g., compare the example ``Last'' of Fig.~\ref{fig:bench-nondet})
to put conditions one computed results.
Solving such conditions by generating all values is not always
a reasonable approach. For instance, if \code{xs} and \code{ys}
are free variables of type \code{[Bool]}, the equational
constraint \ccode{xs=:=ys} has an infinite number of solutions.
Instead of enumerating all these solutions, it is preferable to
delay this enumeration but remember the condition that both
\code{xs} and \code{ys} must always be evaluated to the same value.
This demands for extending the representation of nondeterministic
values by the possibility to add equational constraints between
different choice identifiers.
Due to lack of space, we have to omit the detailed description
of this extension.
However, it should be noted that the example
``Last'' of Fig.~\ref{fig:bench-nondet}
shows that unification can be supported with a reasonable efficiency.

@@

We have presented a new system to compile functional logic programs
into purely functional programs.
In order to be consistent with the call-time choice semantics
of functional logic languages like Curry or Toy,
we represent nondeterministic values
in choice structures where each choice has an identification.
Values for such choice identifiers are passed through
nondeterministic operations so that fresh identifiers are available
when a new choice needs to be created.
The theoretical justification of this implementation technique
is provided in \cite{Brassel11Thesis}.
Apart from the parser, where we reused an existing parser
implemented in Haskell, the compiler is completely written in Curry.

Due to the representation of nondeterministic values as data,
our system easily supports various search strategies
in constrast to Prolog-based implementations.
Since we compile Curry programs into Haskell,
we can exploit the implementation efforts done
for functional programming.
Hence, purely functional parts of functional logic programs
can be executed with almost the same efficiency as Haskell programs.
Our benchmarks show that even the execution of the
nondeterministic parts can compete with other
implementations of Curry.
Furthermore, we can also exploit advanced developments
in the implementation of Haskell, like the parallel evaluation of
programs, which is an interesting topic for future work.
Finally, our implementation has many opportunities
for optimization, like
better program analyses to approximate purely deterministic computations.
